### 2016-09-23 paper/4_gilbert_runtime.tex
Further changes up to middle of evaluation.

The Gilbert runtime is responsible for executing the compiled MATLAB code on a particular platform. For this purpose, it receives the intermediate representation of the code and translates it into the execution engine's specific format. Currently, Gilbert supports three different execution engines: \emph{ReferenceExecutor}, \emph{StratosphereExecutor} and \emph{SparkExecutor}. The \emph{ReferenceExecutor} executes the compiled MATLAB code locally. For the distributed execution Gilbert supports the Spark and the Stratosphere system as backends. The \emph{ReferenceExecutor} is an interpreter for the intermediate code.
It takes the dependency tree of a MATLAB program and executes it by evaluating the operators bottom-up. In order to evaluate an operator, the system first evaluates all inputs of an operator and then executes the actual operator logic. Since the program is executed locally, the complete data of each operator is always accessible and, thus, no communication logic is required. All input and output files are directly written to the local hard drive. In contrast to the \emph{ReferenceExecutor}, the \emph{StratosphereExecutor} executes the program in parallel. It takes the dependency tree of a MATLAB program and translates it into a PACT plan. After the plan is generated, it is issued to the Stratosphere system for parallel execution. This approach implies that the program is not directly executed by the executor. Instead, the executor represents just another translation step. The PACT plans are executed in parallel. Consequently, data structures are needed which can be distributed across several nodes and represent the commonly used linear algebra abstractions, such as vectors and matrices. Moreover, the linear algebra operations have to be adjusted so that they keep working in a distributed environment. Fortunately, Stratosphere offers a rich and expressive API to easily realize distributed computations. The details of the distributed data structures and operations are explained in \cref{sec:DistributedMatrixRepresentation} and \cref{sec:LinearAlgebraOperations}. The \emph{SparkExecutor} is the second executor for distributed computations. In contrast to the \emph{StratosphereExecutor}, it executes the MATLAB programs on top of the Spark system. Since Spark and Stratosphere offer a similar set of programming primitives, they can both operate on the same data structures. Furthermore, even most of the linear algebra operations can be implemented similarly.

@@

In the following, we will outline the implementation behind the matrix multiplication operator (which is most critical for performance in linear algebra programs) in Gilbert as an example. For a reference of the implementation of the remaining operators, we refer the interested reader to~\cite{Rohrmann2014}. Note that we use the Breeze~\cite{breeze} library for conducting local in-operator linear algebraic operations. In a MapReduce-like system there exist two distinct matrix multiplication implementations for square blocking. The first approach is based on replicating rows and columns of the operands and is called replication based matrix multiplication (RMM). The other method is derived from the outer product formulation of matrix multiplications. It is called cross product based matrix multiplication (CPMM). The RMM and CPMM methods have been implemented within SystemML~\cite{ghoting:2011a}. Let us assume that we want to calculate $A \times B = C$ with $A,B$ and $C$ being matrices. The block size has been chosen such that $A$ is partitioned into $m\times l$ blocks, $B$ is partitioned into $l \times n$ blocks and the result matrix $C$ will be partitioned into $m\times n$ blocks. In order to reference the block in the $i$th block row and $j$th block column of $A$, we will write $A_{ij}$. A block row will be denoted by a single subscript index and a block column by a single superscript index. For example, $A_i$ marks the $i$th block row of $A$ and $A^j$ the $j$th block column of $A$. The replication-based strategy will copy for each $C_{ij}$ the $i$th block row of $A$ and the $j$th block column of $B$. The replicated blocks of $A_i$ and $B^j$ will be grouped together. These steps can be achieved by a single mapper. Once this grouping is done, the final result $C_{ij}$ can be calculated with a single reducer. The reduce function simply has to calculate the scalar product of $A_i$ and $B^j$. It is important to stress that $A_i$ is replicated for each $C_{ik},\forall k$. The whole process is illustrated in \cref{fig:RMM}.

@@

In contrast to RMM, CPMM calculates the outer products between $A^k$ and $B_k$ for all $k$. A mapper can group the $A^k$ and $B_k$ together so that a reducer can compute the outer products. Consequently, this method does not replicate any data. The outer product produces intermediate result matrices $C^k$ which have to be added up to produce the final result $C=\sum_{k=1}^{l} C^k$. This summation can be achieved by a subsequent reducer. The whole process is illustrated in \cref{fig:CPMM}. The methods RMM and CPMM differ in terms of network communication. The former method can be realized within a single MapReduce job whereas the latter requires two. Neither RMM nor CPMM is always superior. The optimal matrix multiplication strategy depends on the matrix size of its operands $A$ and $B$. Fortunately, Stratosphere and Spark exhibit a little bit more flexibility in terms of higher order functions. Freed from the tight corset of the MapReduce world, matrix multiplications can be expressed more subtly. Looking at the definition of the matrix multiplication for $C_{ij}=\sum_{k=1}^{l}A_{ik}\times B_{kj}$, it can be seen that every $A_{ik}$ has to be joined with its corresponding $B_{kj},\forall k$. This pairwise mapping can be easily achieved by using the join function. The join-key is the column index of $A$ and the row index of $B$. The joiner calculates for each matching pair $A_{ik}$ and $B_{kj}$ an intermediate result $C_{ij}^k$. Grouping the intermediate results with respect to the index pair $(i,j)$ allows us to compute the final result in a subsequent reduce step. The overall algorithm strongly resembles the CPMM. Stratosphere supports several execution strategies for the higher-order functions. A cost-based optimizer selects the best strategies prior to execution. One possible optimization concerns the join function. The join can either be realized using a hybrid-hash join or a sort-merge join algorithm depending on the current partitioning and the input data sizes. If one input data is relatively small compared to the other input, it is usually more efficient to use the hybrid-hash join algorithm. Without loss of generality, we assume that the matrix $B$ constitutes such a small input. If we further assume that the block rows of $A$ are kept on the same worker node, then the last reduce operation can be executed locally and without any shuffling. The resulting execution plan under these assumptions is equivalent to the RMM. If the system chooses the sort-merge join algorithm instead, then the columns of $A$ will be distributed across the worker nodes. Consequently, the last reduce step causes a mandatory repartitioning. Then, the resulting execution plan is equivalent to the CPMM. Even though Gilbert has only one dataflow plan specified to implement the matrix multiplication, the Stratosphere system can choose internally between the RMM and CPMM strategy. The strategy is selected by the optimizer which bases its decision on the data size and the partitioning, if available. Consequently, Stratosphere frees the programmer from this decision.

>>>>

The Gilbert runtime is responsible for executing the compiled MATLAB code on a particular platform. For this purpose, it receives the intermediate representation of the code and translates it into the execution engine's specific format. Currently, Gilbert supports three different execution engines: \emph{ReferenceExecutor}, \emph{FlinkExecutor} and \emph{SparkExecutor}. The \emph{ReferenceExecutor} executes the compiled MATLAB code locally. For the distributed execution Gilbert supports the Spark and the Flink system as backends. The \emph{ReferenceExecutor} is an interpreter for the intermediate code.
It takes the dependency tree of a MATLAB program and executes it by evaluating the operators bottom-up. In order to evaluate an operator, the system first evaluates all inputs of an operator and then executes the actual operator logic. Since the program is executed locally, the complete data of each operator is always accessible and, thus, no communication logic is required. All input and output files are directly written to the local hard drive. In contrast to the \emph{ReferenceExecutor}, the \emph{FlinkExecutor} executes the program in parallel. It takes the dependency tree of a MATLAB program and translates it into a PACT plan. After the plan is generated, it is issued to the Flink system for parallel execution. This approach implies that the program is not directly executed by the executor. Instead, the executor represents just another translation step. The PACT plans are executed in parallel. Consequently, data structures are needed which can be distributed across several nodes and represent the commonly used linear algebra abstractions, such as vectors and matrices. Moreover, the linear algebra operations have to be adjusted so that they keep working in a distributed environment. Fortunately, Flink offers a rich and expressive API to easily realize distributed computations. The details of the distributed data structures and operations are explained in \cref{sec:DistributedMatrixRepresentation} and \cref{sec:LinearAlgebraOperations}. The \emph{SparkExecutor} is the second executor for distributed computations. In contrast to the \emph{FlinkExecutor}, it executes the MATLAB programs on top of the Spark system. Since Spark and Flink offer a similar set of programming primitives, they can both operate on the same data structures. Furthermore, even most of the linear algebra operations can be implemented similarly.

@@

In the following, we will outline the implementation of the matrix multiplication operator (which is most critical for performance in linear algebra programs) in Gilbert as an example. For a reference of the implementation of the remaining operators, we refer the interested reader to~\cite{Rohrmann2014}. Note that we use the Breeze~\cite{breeze} library for conducting local in-operator linear algebraic operations. In a MapReduce-like system there exist two distinct matrix multiplication implementations for square blocking. The first approach is based on replicating rows and columns of the operands and is called replication based matrix multiplication (RMM). The other method is derived from the outer product formulation of matrix multiplications. It is called cross product based matrix multiplication (CPMM). The RMM and CPMM methods have been implemented within SystemML~\cite{ghoting:2011a}. Let us assume that we want to calculate $A \times B = C$ with $A,B$ and $C$ being matrices. The block size has been chosen such that $A$ is partitioned into $m\times l$ blocks, $B$ is partitioned into $l \times n$ blocks and the result matrix $C$ will be partitioned into $m\times n$ blocks. In order to reference the block in the $i$th block row and $j$th block column of $A$, we will write $A_{ij}$. A block row will be denoted by a single subscript index and a block column by a single superscript index. For example, $A_i$ marks the $i$th block row of $A$ and $A^j$ the $j$th block column of $A$. The replication-based strategy will copy for each $C_{ij}$ the $i$th block row of $A$ and the $j$th block column of $B$. The replicated blocks of $A_i$ and $B^j$ will be grouped together. These steps can be achieved by a single mapper. Once this grouping is done, the final result $C_{ij}$ can be calculated with a single reducer. The reduce function simply has to calculate the scalar product of $A_i$ and $B^j$. It is important to stress that $A_i$ is replicated for each $C_{ik},\forall k$. The whole process is illustrated in \cref{fig:RMM}.

@@

In contrast to RMM, CPMM calculates the outer products between $A^k$ and $B_k$ for all $k$. A mapper can group the $A^k$ and $B_k$ together so that a reducer can compute the outer products. Consequently, this method does not replicate any data. The outer product produces intermediate result matrices $C^k$ which have to be added up to produce the final result $C=\sum_{k=1}^{l} C^k$. This summation can be achieved by a subsequent reducer. The whole process is illustrated in \cref{fig:CPMM}. The methods RMM and CPMM differ in terms of network communication. The former method can be realized within a single MapReduce job whereas the latter requires two. Neither RMM nor CPMM is always superior. The optimal matrix multiplication strategy depends on the matrix size of its operands $A$ and $B$. Fortunately, Flink and Spark exhibit a little bit more flexibility in terms of higher order functions. Looking at the definition of the matrix multiplication for $C_{ij}=\sum_{k=1}^{l}A_{ik}\times B_{kj}$, it can be seen that every $A_{ik}$ has to be joined with its corresponding $B_{kj},\forall k$. This pairwise mapping can be easily achieved by using the join function. The join-key is the column index of $A$ and the row index of $B$. The joiner calculates for each matching pair $A_{ik}$ and $B_{kj}$ an intermediate result $C_{ij}^k$. Grouping the intermediate results with respect to the index pair $(i,j)$ allows us to compute the final result in a subsequent reduce step. The overall algorithm strongly resembles the CPMM. Flink supports several execution strategies for the higher-order functions. A cost-based optimizer selects the best strategies prior to execution. One possible optimization concerns the join function. The join can either be realized using a hybrid-hash join or a sort-merge join algorithm depending on the current partitioning and the input data sizes. If one input data is relatively small compared to the other input, it is usually more efficient to use the hybrid-hash join algorithm. Without loss of generality, we assume that the matrix $B$ constitutes such a small input. If we further assume that the block rows of $A$ are kept on the same worker node, then the last reduce operation can be executed locally and without any shuffling. The resulting execution plan under these assumptions is equivalent to the RMM. If the system chooses the sort-merge join algorithm instead, then the columns of $A$ will be distributed across the worker nodes. Consequently, the last reduce step causes a mandatory repartitioning. Then, the resulting execution plan is equivalent to the CPMM. Even though Gilbert has only one dataflow plan specified to implement the matrix multiplication, the Flink system can choose internally between the RMM and CPMM strategy. The strategy is selected by the optimizer which bases its decision on the data size and the partitioning, if available. %Consequently, Flink frees the programmer from this decision.
